Distortion corrected kernel density estimate on Riemannian manifolds

### 2022-11-07 paper/paper.Rmd
final polishing of 1 and 2

# Introduction

@@

$$Pr(\pmb{y}\in\psi(\mathcal{A}))=\int_{\psi(\mathcal{A})} f(\psi^{-1}(\pmb{y}))|g(\pmb{y})|^{-1/2}d\pmb{y}$$
where $\psi(\mathcal{A})$ is the image of $\mathcal{A}$ under $\psi$ and $g(\pmb{y})$ is the Riemannian metric expressed in local coordinates given by the mapping $\psi$. It can be seen that the density in terms of $\pmb{y}$, differs from the density on the manifold due to the term $|g(\pmb{y})|^{-1/2}$ which can be considered roughly analogous to the Jacobian determinant. Since $\psi$ will typically be a non-linear function, the value of $|g(\pmb{y})|^{-1/2}$ will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm. Notably, different choices of dimension reduction algorithm, imply different $\psi$, meaning that if KDE is applied to the output $\pmb{y}$ coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.

In this paper we propose a new KDE. Our estimator is also based on the output of dimension reduction algorithms, but accounts for the distortion induced by the algorithm. For this reason we refer to it as the distortion corrected KDE (DC-KDE). Critical to our estimator, is obtaining an estimate of $|g(\pmb{y})|^{-1/2}$, for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point. One particular downstream task that we are interested in is that of anomaly detection, where anomalies are identified as points of low density. In a number of simulation studies we demonstrate two benefits that arise when using our new estimator. The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used. The second is that when the estimates are found using different dimension reduction algorithms (e.g. ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.

@@

In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding. Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of @Pelletier2005-vu, when the data lie on some known manifold. In \autoref{MetLearn}, we describe the Learn Metric algorithm of @Perrault-Joncas2013-pq, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates. By combining elements from the work of @Pelletier2005-vu and @Perrault-Joncas2013-pq, we derive our own novel distortion corrected kernel density estimate in \autoref{DCKDE}. To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.

In the following, we denote $M$ as the $d$-dimensional manifold from which our data are sampled. Points on this manifold are denoted $\pmb{p}$ in general, with $\pmb{p}_1,\dots,\pmb{p}_n$ denoting the observed sample. Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m\gg d$, however this need not be the case. For instance, $\pmb{p}_i$ may be probability distributions on a statistical manifold. The methods we propose for estimating the density at each $\pmb{p}_i$ only require some sense of distance between the 'input' points, $d(\pmb{p}_i,\pmb{p}_j)$, such that we can apply dimension reduction algorithms to obtain an `output' embedding, $\pmb{y}_1,\dots,\pmb{y}_n$, where $\pmb{y}_i\in\mathbb{R}^d$. We will denote this embedding as $\pmb{y}_i=\psi(\pmb{p}_i)$. Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $\|\cdot\|$ be the usual Euclidean norm and following @Pelletier2005-vu, we make these assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$,
\begin{align}
\label{eq:kernelcondition}
& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=1;
(ii) \int_{\mathbb{R}^{d}} \pmb{y} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=0;
(iii) \int_{\mathbb{R}^{d}}\|\pmb{y}\|^{2} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})<\infty; \\
& (iv) \supp K=[0; 1];
(v) \sup K(\|\pmb{y}\|)=K(0).
\end{align}
Note that these conditions are different from (and in some cases stricter than) those normally used for kernel density estimation. For instance, condition (iv) requires the support of the kernel to be bounded. The reasons for this will become clearer when we discuss the manifold setting in more detail. Also, for illustration purposes, in this section we pay particular attention to the uniform kernel for which $K(z)$ equals one if $0\leq z\leq1$ and zero otherwise. In our empirical section, more general kernel functions can be, and are, employed.

@@

The intuition behind this estimator is very clear for a uniform kernel. The density at a point $\pmb{y}$ is equal to the proportion of sample points that lie within a ball of radius $r$ centered at $\pmb{y}$, times a term that ensures the density integrates to 1. In general, the bandwidth matrix need not be proportional to the identity matrix. However, the intuition remains the same, only that the ball of radius $r$ centered at $\pmb{y}$ is found with respect to Mahalanobis distance rather than the usual Euclidean distance. For more on kernel density estimation in the Euclidean case, see @Scott2015-vl and references therein.

Kernel density estimators of this form can and have been applied directly on the output embedding $\pmb{y}$, and we will consider this approach as a benchmark in \autoref{simulation}. As a non-linear transformation, any dimension reduction algorithm `distorts' the density. To make this clear consider the simpler case computing the density after a change of variables $\psi:\mathbb{R}^d\rightarrow\mathbb{R}^d$, which involves a Jacobian term. A similar notion applies to a manifold embedding so that the density of the output vectors $\pmb{y}_i$ differs from the density on the manifold itself.  Furthermore, standard kernel density estimates applied directly on the output embedding will be sensitive to the choice of dimension reduction algorithm since each different algorithm will distort the density in its own. This motivates a kernel desnity estimate that corrects for the distortion induced by $\psi$.

@@

The inclusion of the volume density function  $\theta_{\pmb{p}_i}(\pmb{p})$ is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind the inclusion of this term. We have already highlighted that when using a uniform kernel, the kernel density estimate at a point $\pmb{p}$ directly depends on the proportion of sample points within a ball of radius $r$ around $\pmb{p}$. However, the volume of this ball must also be taken into account. In Euclidean space with the usual Lebesgue measure, a radius $r$ ball will always have the same volume regardless of its center. The same does not hold for manifolds and the volume density function ensures that the density estimate integrates to one.

More formally, the volume density function can be explained as follows. Consider the exponential map around $\pmb{p}$, given by $exp_{\pmb{p}}(\pmb{q})$, mapping vectors in the tangent space, $\pmb{v}\in T_{\pmb{p}}M$, to points on the manifold, $\pmb{q}\in M$. Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and travelling along this geodesic at uniform speed $\|\pmb{v}\|$ takes place in one unit of time. Now, consider a chart $\varphi$ mapping points in the neighborhood of $\pmb{p}$, via the inverse of the exponential map, to these $\pmb{v}$ vectors, expressed in some local coordinate system. The volume density function is the square root of the determinant of the Riemannian metric expressed in this coordinate system. For more on the volume density function, see @Le_Brigant2019-lj.

@@

To be able to apply the estimator of @Pelletier2005-vu to the case where the manifold is not known, but where coordinates $\pmb{y}_i$ for $i=1,\dots,n$ are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric in the coordinate system. Formally, the Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $\langle,\rangle_g$ on the tangent space $T_{\pmb{p}} M$ for every point $\pmb{p} \in M$. The inner product between two tangent vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g$, can be used to define geometric quantities. For example, angles on a manifold are given by $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$, while distances and volumes on manifolds are also defined with reference to the Riemannian metric. While the defined tangent vectors, the Riemannian metric and the geometric quantities are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinate systems. This is precisely the situation when data on a manifold are mapped to $d$-dimensional Euclidean vectors $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ via a dimension reduction algorithm. After this mapping, angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions. To alleviate this issue, @Perrault-Joncas2013-pq propose a method to augment $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ with $d\times d$ positive definite matrices, $\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n$, at each data point. These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm. For example, the angle between $\pmb{p}_j$ and $\pmb{p}_k$ at $\pmb{p}_i$ depends (up to a first order approximation) on the inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}^{-1}_i(\pmb{y}_k-\pmb{y}_i)$ rather than the usual Euclidean inner product $(\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)$.

>>>>

# Introduction{#intro}

@@

\begin{equation}
\label{eq:changevar}
Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y}
\end{equation}

where $\psi(\mathcal{A})$ is the image of $\mathcal{A}$ under $\psi$ and $\pmb{H}(\pmb{y})$ is the Riemannian metric expressed in local coordinates given by the mapping $\psi$. It can be seen that the density in terms of $\pmb{y}$, differs from the density on the manifold due to the term $|\pmb{H}(\pmb{y})|^{1/2}$ which can be considered roughly analogous to the Jacobian determinant. Since $\psi$ will typically be a non-linear function, the value of $|\pmb{H}(\pmb{y})|^{1/2}$ will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm. Notably, different choices of dimension reduction algorithm, imply different $\psi$, meaning that if KDE is applied to the output $\pmb{y}$ coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.

In this paper we propose a new KDE. Our estimator is also based on the output of dimension reduction algorithms, but accounts for the distortion induced by the algorithm. For this reason we refer to it as the distortion corrected KDE (DC-KDE). Critical to our estimator, is obtaining an estimate of $|\pmb{H}(\pmb{y})|^{1/2}$, for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point. One particular downstream task that we are interested in is that of anomaly detection, where anomalies are identified as points of low density. In a number of simulation studies we demonstrate two benefits that arise when using our new estimator. The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used. The second is that when the estimates are found using different dimension reduction algorithms (e.g. ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.

@@

In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding. Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of @Pelletier2005-vu, which can be applied when the data lie on some known manifold. In \autoref{MetLearn}, we describe the Learn Metric algorithm of @Perrault-Joncas2013-pq, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates. By combining elements from the work of @Pelletier2005-vu and @Perrault-Joncas2013-pq, we derive our own novel distortion corrected kernel density estimate in \autoref{DCKDE}. To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.

In the following, we denote $M$ as the $d$-dimensional manifold from which our data are sampled. Points on this manifold are denoted $\pmb{p}$ in general, with $\pmb{p}_1,\dots,\pmb{p}_n$ denoting the observed sample. Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m\gg d$, however this need not be the case. For instance, $\pmb{p}_i$ may be probability distributions on a statistical manifold. The methods we propose for estimating the density at each $\pmb{p}_i$ only require some sense of distance between the 'input' points, $d(\pmb{p}_i,\pmb{p}_j)$, such that we can apply dimension reduction algorithms to obtain an `output' embedding, $\pmb{y}_1,\dots,\pmb{y}_n$, where $\pmb{y}_i\in\mathbb{R}^d$. We will denote this embedding as $\pmb{y}_i=\psi(\pmb{p}_i)$. Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $\|\cdot\|$ be the usual Euclidean norm and following @Pelletier2005-vu, we make the following assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$,
\begin{align}
\label{eq:kernelcondition}
& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=1;
(ii) \int_{\mathbb{R}^{d}} \pmb{y} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=0;
(iii) \int_{\mathbb{R}^{d}}\|\pmb{y}\|^{2} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})<\infty; \\
& (iv) \supp K=[0; 1];
(v) \sup K(\|\pmb{y}\|)=K(0).
\end{align}
Note that these conditions are different from (and in some cases stricter than) those normally used for kernel density estimation. For instance, condition (iv) requires the support of the kernel to be bounded. The reasons for this will become clearer when we discuss the manifold setting in more detail. Also, for illustration purposes, in this section we pay particular attention to the uniform kernel for which $K(z)$ equals one if $0\leq z\leq1$ and zero otherwise. In our empirical section, more general kernel functions can be, and are, employed.

@@

The intuition behind this estimator is very clear for a uniform kernel. The density at a point $\pmb{y}$ is equal to the proportion of sample points that lie within a ball of radius $r$ centered at $\pmb{y}$, times a term that ensures the density integrates to 1. In general, the bandwidth matrix need not be proportional to the identity matrix. However, the intuition remains the same, only that the ball of radius $r$ centered at $\pmb{y}$ is found with respect to Mahalanobis distance rather than the usual Euclidean distance. For more on kernel density estimation in the Euclidean case, see @Scott2015-vl and references therein. Kernel density estimators of this form can and have been applied directly on the output embedding $\pmb{y}$, and we will consider this approach as a benchmark in \autoref{simulation}.

<!--As a non-linear transformation, any dimension reduction algorithm `distorts' the density. To make this clear consider the simpler case computing the density after a change of variables $\psi:\mathbb{R}^d\rightarrow\mathbb{R}^d$, which involves a Jacobian term. A similar notion applies to a manifold embedding so that the density of the output vectors $\pmb{y}_i$ differs from the density on the manifold itself.  Furthermore, standard kernel density estimates applied directly on the output embedding will be sensitive to the choice of dimension reduction algorithm since each different algorithm will distort the density in its own. This motivates a kernel desnity estimate that corrects for the distortion induced by $\psi$.-->

@@

The inclusion of the volume density function  $\theta_{\pmb{p}_i}(\pmb{p})$ is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind the inclusion of this term. We have already highlighted that when using a uniform kernel, the kernel density estimate at a point $\pmb{p}$ directly depends on the proportion of sample points within a ball of radius $r$ around $\pmb{p}$. However, the volume of this ball must also be taken into account. In Euclidean space with the usual Lebesgue measure, a radius $r$ ball will always have the same volume regardless of its center. The same does not hold for manifolds and therefore the volume density function must be included in the KDE to ensure that the estimated density integrates to one.

More formally, the volume density function can be explained as follows. Consider the exponential map around $\pmb{p}$, given by $exp_{\pmb{p}}(\pmb{q})$, mapping vectors in the tangent space,  $\pmb{v}\in T_{\pmb{p}}M$, to points on the manifold, $\pmb{q}\in M$. Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and travelling along this geodesic at uniform speed $\|\pmb{v}\|$ takes place in one unit of time. Now, consider a chart $\varphi$ mapping points in the neighborhood of $\pmb{p}$, via the inverse of the exponential map, to these $\pmb{v}$ vectors, expressed in some local coordinate system. The volume density function is then the square root of the determinant of the Riemannian metric expressed in this coordinate system. In proving that the estimator integrates to 1, the volume density function cancels with the term $|\pmb{H}(\pmb{y})|^{1/2}$ in Equation \@ref(eq:changevar), where $\psi$ is the logarithmic map around $\pmb{p}_i$. For more on the volume density function, see @Le_Brigant2019-lj.

@@

To be able to apply the estimator of @Pelletier2005-vu to the case where the manifold is not known, but where coordinates $\pmb{y}_i$ for $i=1,\dots,n$ are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric. Formally, the Riemannian metric is a symmetric and positive definite tensor field which defines an inner product on the tangent space $T_{\pmb{p}} M$ for every point $\pmb{p} \in M$. <!--The inner product between two tangent vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g$, can be used to define geometric quantities. For example, angles on a manifold are given by $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$, while distances and volumes on manifolds are also defined with reference to the Riemannian metric. While the defined tangent vectors, the Riemannian metric and the geometric quantities are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinate systems. This is precisely the situation when data on a manifold are mapped to $d$-dimensional Euclidean vectors $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ via a dimension reduction algorithm.--> Using the coordinates from a dimension reduction algorithm $\pmb{y}$, angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions. To alleviate this issue, @Perrault-Joncas2013-pq propose a method to augment $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ with $d\times d$ positive definite matrices, $\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n$, at each data point. These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm. <!--For example, the angle between $\pmb{p}_j$ and $\pmb{p}_k$ at $\pmb{p}_i$ depends (up to a first order approximation) on the inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}^{-1}_i(\pmb{y}_k-\pmb{y}_i)$ rather than the usual Euclidean inner product $(\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)$. -->
